---
title: "meta_analysis"
author: "Haris Babačić"
date: "10/03/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r load_packages, include=FALSE}
pack <- c("tidyverse", "ggplot2", "matrixStats", "VennDiagram", "ggrepel", "dplyr", "factoextra", "FactoMineR", "ggfortify", "cluster", "gplots", "RColorBrewer", "ComplexHeatmap", "readxl", "cutpointr", "pROC")
lapply(pack, require, character.only = TRUE)

```

Functions for calculating input data for the meta analysis
```{r functions_for_meta_analysis, include=FALSE}
### Function for calculating mean, number of participants in which a protein has been quantified, and standard deviation
summary_func = function(pt, hc, sdf) {
  groups = list(pt, hc)
  names(groups) = c("pt", "hc")
  mean = list()
  sd = list()
  number = list()
  
  for (i in names(groups)) {
  sel = which(colnames(sdf) %in% groups[[i]])
  mean[[i]] <- apply(sdf[,sel], 1, function(x) mean(x, na.rm=T))
  number[[i]] <- apply(sdf[,sel], 1, function(x) length(which(!is.na(x))))
  sd[[i]] <- apply(sdf[,sel], 1, function(x) sd(x, na.rm=T))
  }
  
  sum <- as.data.frame(cbind(do.call(cbind, mean), do.call(cbind, sd), do.call(cbind, number)))
  colnames(sum) <- c(paste("mean", c("covid", "healthy"), sep="_"), 
                  paste("sd", c("covid", "healthy"), sep="_"),
                  paste("number", c("covid", "healthy"), sep="_"))
  sum
}

### Function for calculating ROC curves per protein in a study, and estimating optimal cut-off point
rocs_func = function(x, selprot, file, plot) {
  x = as.data.frame(x)
  
  for (i in selprot) {
  cpp[[i]] <<- cutpointr(x[,i], x$type, pos_class = "COVID19", 
               method = maximize_metric, metric = youden, na.rm=T)
  tp[[i]] <<- cpp[[i]]$roc_curve[[1]]
  
  if(cpp[[i]]$direction=="<=") {
    t[[i]] <<- as.vector(table(x[,i]<=cpp[[i]]$optimal_cutpoint, x$type))
    names(t[[i]]) <<- c("fn", "tp", "tn", "fp")
  } else {
    t[[i]] <<- as.vector(table(x[,i]>=cpp[[i]]$optimal_cutpoint, x$type))
    names(t[[i]]) <<- c("fn", "tp", "tn", "fp")
  }
  
  if (plot==TRUE) {
    tr_roc = ggplot(tp[[i]], aes(fpr, tpr)) +
    geom_line(col="red", size=1) +
    theme_classic() +
    geom_abline(intercept = 0, slope=1, linetype="dashed", col="orange", size=0.75) +
    scale_y_continuous("Sensitivity (TPR)", limits=c(0,1.02), 
                       breaks=seq(0,1,0.2)) +
    scale_x_continuous("1-Specificity (FPR)", breaks=seq(0,1,0.2)) +
    ggtitle(i) +
    geom_point(x=1-cpp[[i]]$specificity, y=cpp[[i]]$sensitivity, color="navyblue", size=3) +
    geom_text(x=1-cpp[[i]]$specificity, y=cpp[[i]]$sensitivity+0.05, 
              label=round(cpp[[i]]$optimal_cutpoint, 3))
  
  ggsave(paste("~/Desktop/covid_hirief/results/roc_curves/", file, "/", i, ".pdf", sep=""), 
         tr_roc, dpi=600, width=3.5, height=3.5)
  }
}
}

### Function to save the data
save_runs = function(i) {
  tp_summary[[i]] <<- tp
  t_summary[[i]] <<- do.call(rbind, t)
  cpp_summary[[i]] <<- as.data.frame(do.call(rbind, cpp))
  sdf_files[[i]] <<- sdf
  sume_files[[i]] <<- sum
}

### Function for a modified run in instances where the protein has been quantified several times
# In those instances, we used the median value of the protein quants
sdf2_func = function(sdf, genenames, protnames) {
  
  dupl = duplicated(genenames)
  sel = which(dupl==T)
  chosen1 = unique(genenames[sel])
  rem = c(which(is.na(chosen1)), which(chosen1==""))
  chosen = chosen1[-rem]
  
  sdfs = list()
  for(z in chosen) {
    sel = which(genenames==z)
    sdfss = as.matrix(sdf[sel,])
    sdfss = apply(sdfss, 2, as.numeric)
    sdfs[[z]] = apply(sdfss, 2, function(x) {median(x, na.rm=T)})
  }
  
  sdf = sdf[-which(genenames %in% chosen),]
  sdf2 = rbind(sdf, do.call(rbind, sdfs))
  
  rownames(sdf2)[match(chosen, rownames(sdf2))] = protnames[match(chosen, genenames)]
  
  sdf2
}

### Function to calculate missing values
miss_func = function(x) {
  apply(x, 2, function(x) length(which(is.na(x)))/length(x)*100)
}

```

Create empty list to fill in with the data from the different studies
```{r create_roc_lists, include=FALSE} 
sdf_files = list()
tp_summary = list()
cpp_summary = list()
t_summary = list()
sume_files = list()

```

An example of using the functions on our dataset
```{r summary_ourstudy, include=FALSE}
# Don't forget to set the right working directory!
setwd("code_submission/data")
# Load Uniprot annotatio file with different gene IDs
annot = read.csv("protein_annotation_file.csv")
# Load a list of proteins with at least 3 observations in a study
prot_suff = read_rds("proteins_with_at_least_3_obs.RData")

data = as.data.frame(read.delim("genes_table_serum.txt"))
sdf = as.matrix(data[,15:46])
sdf = sdf[,-grep("IS", colnames(sdf))]
rownames(sdf) = data$Gene.Name

pt = colnames(sdf)[-grep("HC", colnames(sdf))]
hc = colnames(sdf)[grep("HC", colnames(sdf))]

sum = summary_func(pt, hc, sdf)

sum$gene_name = rownames(sum)
sum$uniprot = annot$uniprotswissprot[match(sum$gene_name, annot$external_gene_name)]
sum$gene_id = annot$ensembl_gene_id[match(sum$gene_name, annot$external_gene_name)]
sum$prot_id = annot$ensembl_peptide_id[match(sum$gene_name, annot$external_gene_name)]
sum$entrez_id = annot$entrezgene_id[match(sum$gene_name, annot$external_gene_name)]

write.csv(sum, "../results/processed_data/summary_estimates/babacic_2022_summary.csv", row.names = F)

#### ROC curves
x = as.data.frame(t(sdf))
#miss = apply(x, 2, function(x) length(which(is.na(x)))/length(x))
#rem = which(miss>0.5)
sel = which(colnames(x) %in% prot_suff[[grep("baba", names(prot_suff))]])
x = x[,sel]
x$type = "HC"
x$type[match(pt, rownames(x))] = "COVID19"
selprot = colnames(x)[-ncol(x)]

tp = list()
cpp = list()
t = list()

rocs_func(x, selprot, "babacic_2022", FALSE)
save_runs(1)

```

Meta-analysis for SMD
```{r meta_analysis, include=FALSE}
# Calculate s_pooled
library(esc)
library(dmetar)
library(meta)
library(tidyverse)

setwd("../results/processed_data/summary_estimates")
files = list.files()
files

ns = length(files)

df = list()
for(i in 1:ns) {
  x = read.csv(files[i])
  rem1 = which(x$number_covid<3)
  rem2 = which(x$number_healthy<3)
  rem = union(rem1, rem2)
  if(is_empty(rem)) {
    df[[i]] = x
  } else {
    df[[i]] = x[-rem,]
  }
}

#View(df)

prot_suff = list()
for (i in 1:ns) {
  prot_suff[[i]] = df[[i]]$gene_name
}
names(prot_suff) = files
write_rds(prot_suff, "../../../data/proteins_with_at_least_3_obs.RData")

names = list()
names[[1]] = df[[1]]$gene_name
for(i in 2:ns) {
  names[[i]] = union(df[[i]]$gene_name, names[[i-1]])
}

det_prot = as.data.frame(names[[i]])
colnames(det_prot) = "gene_name"

for (i in 1:ns) {
  sel = which(det_prot$gene_name %in% df[[i]]$gene_name)
  det_prot[,i+1] = 0
  det_prot[,i+1][sel] = 1
}

colnames(det_prot)[2:(ns+1)] = str_to_title(gsub("_.+", "", files))

det_prot$nstud = apply(det_prot[,2:(ns+1)], 1, sum)
#View(det_prot)
table(det_prot$nstud)
rem = c(which(is.na(det_prot$gene_name)), which(det_prot$gene_name==""))
det_prot = det_prot[-rem,]
dbm_det_prot = det_prot[grep(";", det_prot$gene_name),]
length(grep(";", det_prot$gene_name))/nrow(det_prot)*100
det_prot = det_prot[-which(det_prot$gene_name %in% dbm_det_prot$gene_name),]

mg.sum = list()
sume = list()
sume_calc = list()
cases = list()
controls = list()
nstud = list()
mg.selected = c(13:31, 33, 35:44, 47:56,60:62, 74, 81)

# Select the intersecting proteins
proteins = det_prot$gene_name[which(det_prot$nstud>1)] # 35 proteins had matched to several genes

for (i in 1:length(proteins)) {
  prot = proteins[[i]]
  param = list()
  eff_size = list()
  g_eff = list()
  
  for (z in 1:ns){
  x = df[[z]]
  param[[z]] = x[which(x$gene_name==prot),1:6]
  names(param)[[z]] = files[[z]]
  }
  
  sume = as.data.frame(do.call(rbind, param))
  
    sume$author = gsub("_.+", "", rownames(sume))
    sume$author = paste(str_to_title(sume$author), "et al.")
    sume$publication_year = str_extract(rownames(sume), "[0-9]+")
    
    sume_calc <- esc_mean_sd(grp1m = sume$mean_covid,
                       grp1sd = sume$sd_covid,
                       grp1n = sume$number_covid,
                       grp2m = sume$mean_healthy,
                       grp2sd = sume$sd_healthy,
                       grp2n = sume$number_healthy,
                       study = sume$author,
                       es.type = "g") %>% 
                     as.data.frame()
    
    sume_calc$w <- 1/sume_calc$se^2
    
    cases[[i]] = sum(sume$number_covid)
    controls[[i]] = sum(sume$number_healthy)
    nstud[[i]] = length(sume$author)
    
    m.gen <- metagen(TE = es,
                 seTE = se,
                 studlab = study,
                 data = sume_calc,
                 sm = "SMD",
                 fixed = FALSE,
                 random = TRUE,
                 method.tau = "REML",
                 #hakn = TRUE,
                 title = paste("Change in protein:", prot))
    summary(m.gen)
    
    m.gen <- update.meta(m.gen, prediction = TRUE)
    summary(m.gen)
    
    #find.outliers(m.gen)
    #m.gen.inf <- InfluenceAnalysis(m.gen, random = TRUE)
    #plot(m.gen.inf, "baujat") # error
    
    sume[[i]] = sume
    sume_calc[[i]] = sume_calc
    
    names(m.gen)
    
    mg.sum[[i]] = unlist(m.gen[mg.selected])
    
    pdf(file = paste("../forest_plots/", prot, "_forestplot.pdf", sep=""), width = 8, height = 7)
    forest.meta(m.gen, 
            sortvar = TE,
            prediction=TRUE,
            print.tau2 = FALSE,
            leftlabs = c("Author", "n", "g", "SE"))
    dev.off()
    
    print(paste("Run:", i, "done."))
    print(paste("Protein ", proteins[[i]], " identified in: ", nrow(sume), " studies.", sep=""))
    
}

meta_results = as.data.frame(cbind(do.call(rbind, mg.sum), unlist(cases), unlist(controls)))
rem = c(15, 19, 31, 41:45)
meta_results[,-rem] = apply(meta_results[,-rem], 2, function(x) as.numeric(as.character(x)))
rownames(meta_results) = proteins
colnames(meta_results)[46:47] = c("cases", "controls")

meta_results$sum = meta_results$cases+meta_results$controls
meta_results$gene_name = proteins
meta_results = meta_results[,c(49:46, 1:45)]

meta_results$random.adj.pval = p.adjust(meta_results$pval.random, method="fdr")
meta_results$fixed.adj.pval = p.adjust(meta_results$pval.fixed, method="fdr")
sel = which(meta_results$pval.random<0.05)

meta_sel = meta_results[sel,]
table(meta_sel$fixed.adj.pval<0.05)
table(meta_sel$random.adj.pval<0.05)

setwd("../../")

meta_sel = meta_sel[order(meta_sel$TE.random, decreasing = T),]
meta_sel$order = nrow(meta_sel):1

meta_sel2 = meta_sel[which(meta_sel$k.study==11),]
meta_sel2$cat = ifelse(meta_sel2$TE.random>0, "Up in COVID-19", "Down in COVID-19")

#sel = which(meta_sel2$random.adj.pval<0.05)
mp = ggplot(meta_sel2, aes(TE.random, as.factor(order), col=cat)) +
  geom_errorbar(aes(xmin=lower.random, xmax=upper.random)) +
  geom_point(size=3, shape="diamond") +
  scale_y_discrete("Proteins (gene name)", labels=rev(meta_sel2$gene_name)) +
  theme_classic() +
  geom_vline(xintercept = 0, col="black", linetype="dashed", size=1) +
  scale_x_continuous("Random effect model: Estimate + 95% CI", breaks=seq(-5.5, 5, 0.5)) +
  scale_color_manual("Alteration", values=c("#0072B2", "#D55E00")) +
  theme(legend.position = c(0.8, 0.2))
mp

ggsave("meta_analysis_sign_all_forest_11_studies.pdf", mp, width=9, height=7)

intersect(meta_sel$gene_name, det_prot$gene_name[which(det_prot$nstud==ns)])

```

Volcano plot for the meta analysis of SMD
```{r volcano_meta, include=FALSE, echo=FALSE}
tres = meta_results

tres$alteration = "Up in Covid-19"
tres$alteration[which(tres$TE.random<0)] = "Down in Covid-19"

tres$shade = abs(tres$TE.random)*-log10(tres$pval.random)
tres$shadcat = cut(tres$shade, breaks=c(0,0.25, 0.5, 0.75, 1, 1.5, 2.5, 7, 16))
levels(tres$shadcat) = c(1:8)
table(is.na(tres$shadcat))
table(tres$shadcat)

y <- tres[which(tres$shade>2.5 & tres$TE.random>0),]
r <- tres[which(tres$shade>2.5 & tres$TE.random<0),]

cov = ggplot(tres, aes(x=TE.random, y=-log10(pval.random), col=alteration)) + 
    geom_point(aes(alpha=as.numeric(shadcat))) + 
    theme_classic() + 
    ylab(expression(paste("-log10 ", italic("p"), " values"))) + 
    xlab("SMD") + 
    scale_alpha(range=c(0.1,1), guide="none") +
    geom_hline(yintercept=-log10(0.05), col="red", linetype="dotdash", size=1) + 
    scale_color_manual("Alteration", values=c("#0072B2", "#D55E00")) +
    theme(legend.position = c(0.9, 0.9)) +
    geom_text_repel(data=y, aes(x=TE.random, y=-log10(pval.random), label=gene_name), col="#D55E00", vjust=0.5, hjust=0.1) + 
    geom_text_repel(data=r, aes(x=TE.random, y=-log10(pval.random), label=gene_name), col="#0072B2", vjust=0.5, hjust=-0.1) + 
    scale_x_continuous("", breaks=seq(round(min(tres$TE.random)),round(max(tres$TE.random)), 0.5)) +
    scale_y_continuous("", breaks=seq(round(min(-log10(tres$pval.random), digits=1)), 
                                  round(max(-log10(tres$pval.random)), digits=1), 1))
cov

ggsave("volcano_meta_analysis.pdf", cov, width=6.25, height=5, units="in", dpi=600)

quartz(type = 'pdf', file = 'volcano_meta_analysis.pdf', width=6.25, height=5)
cov
dev.off()

```

Comparison between statistically significant proteins in HiRIEF and in the meta analysis
```{r comparison_to_hirief, include=FALSE, echo=FALSE}
# Meta-comparison
tres = read.csv("covid_vs_hc_all_res.csv")

ov = intersect(meta_results$gene_name, tres$gene)
res = tres
res = res[which(res$gene %in% ov),]
res = res[which(res$adj.p<0.05),]

# Change metsign depending on the significance you want
res$metsign = meta_results$TE.random[match(res$gene, meta_results$gene_name)]
res$metsign = meta_sel$TE.random[match(res$gene, meta_sel$gene_name)]
table(is.na(res$metsign))[1]/nrow(res)

res$olcat = NA
res$olcat[which(res$fc>0 & res$metsign>0)] = paste('\U2191', "HiRIEF-MS,", '\U2191', "Meta-analysis")
res$olcat[which(res$fc>0 & res$metsign<0)] = paste('\U2191', "HiRIEF-MS,", '\U2193', "Meta-analysis")
res$olcat[which(res$fc<0 & res$metsign<0)] = paste('\U2193', "HiRIEF-MS,", '\U2193', "Meta-analysis")
res$olcat[which(res$fc<0 & res$metsign>0)] = paste('\U2193', "HiRIEF-MS,", '\U2191', "Meta-analysis")
t = table(res$olcat)
a = sum(t[c(1,2)])/sum(t)*100

cv = cor(res$fc, res$metsign, use="pairwise.complete.obs", method="spearman")

#sel = which(!is.na(res$metsign))
#For all colours: "#D55E00", "#E69F00", "#CC79A7", "#009E73"

p = ggplot(res, aes(fc, metsign, col=olcat)) +
  geom_point() +
  geom_text_repel(label=res$gene) + 
  theme_classic() +
  geom_vline(xintercept = 0, col="darkgray", linetype="dashed", size=0.5) +
  geom_hline(yintercept = 0, col="darkgray", linetype="dashed", size=0.5) +
  scale_x_continuous("HiRIEF LC-MS/MS, log2-FC COVID-19:Healthy", breaks=seq(-2,2,0.25)) +
  scale_y_continuous("Meta-analysis, log2-FC COVID-19:Controls", breaks=seq(-3,3.5,0.5)) +
  annotate("text", x=-1.5, y=1.5, hjust=0, col="black",
           label=paste("Agreement: ", round(a, digits=3), "%", sep="")) +
  annotate("text", x=-1.5, y=1.3, hjust=0, col="black",
           label=paste("r = ", round(cv, digits=3), sep="")) +
  scale_color_manual(name="change", values=c("#D55E00", "#009E73")) +
  theme(legend.position = c(0.8, 0.15))
p

quartz(type = 'pdf', file = 'Meta_analysis_sign_vs_HiRIEF_adjpval_sign.pdf', width=8, height=6)
p
dev.off()
ggsave("Meta_analysis_sign_vs_HiRIEF_adjpval_sign.jpeg", p, dpi=600, width=8, height=6)

```

Meta analysis on SROC curves
```{r meta_analysis_roc_curves, include=FALSE}
library("mada")
library("ggforce")
library(plotly)

t_summary = read_rds("t_summary.RData")

studies = c("Babacic", "Shen", "Geyer", "Shu", "Di", "Messner-Discovery",
            "Messner-Validation", "Suvarna", "Sullivan", "Overmyer", "Zhang")

names = list()
names[[1]] = rownames(t_summary[[1]])
for(i in 2:ns) {
  names[[i]] = union(rownames(t_summary[[i]]), names[[i-1]])
}

roc_prot = as.data.frame(names[[i]])
colnames(roc_prot) = "gene_name"

for (i in 1:ns) {
  sel = which(roc_prot$gene_name %in% rownames(t_summary[[i]]))
  roc_prot[,i+1] = 0
  roc_prot[,i+1][sel] = 1
}
colnames(roc_prot)[-1] = studies
roc_prot$nstud = apply(roc_prot[,-1], 1, sum)

##### Select proteins of interest
selected = roc_prot$gene_name[which(roc_prot$gene_name %in% meta_results$gene_name 
                                    & roc_prot$nstud>1)]
# These have been removed because there were too few observations to make the analysis
selected = selected[-which(selected %in% c("RANBP2", "VSIG4", "ENG", "ASGR2", "CLSTN1", 
                                           "PSMA4", "NCAM2", "SNED1"))]

mtdats = list()
fit.reitsmas = list()
fit.DOR.DSLs = list()
sumfit.DOR = list()
sumfit.reits = list()
coef1 = list()
coef2 = list()
coef = list()

for(z in selected) {
  values = list()
  prot = z
  
  for(i in 1:11) {
    t = as.data.frame(t_summary[[i]])
    if (prot %in% rownames(t)) {
      values[[i]] = t_summary[[i]][prot,]
      } else {
        values[[i]] = NA
      }
    }
  
  mtdat = as.data.frame(do.call(rbind, values))
  colnames(mtdat) = c("FN", "TP", "TN", "FP")
  mtdat = na.omit(mtdat)
  mtdats[[z]] = mtdat
  
  fit.DOR.DSL <- madauni(mtdat, correction=0.5)
  fit.DOR.DSLs[[z]] = fit.DOR.DSL
  
  sumfit.DOR[[z]] = summary(fit.DOR.DSL)
  coef1[[z]] = c(sumfit.DOR[[z]]$CIcoef[1,], sumfit.DOR[[z]]$CIcoef[2,], sumfit.DOR[[z]]$CIcoef[3,], sumfit.DOR[[z]]$CIcoef[4,],
              sumfit.DOR[[z]]$Isq, sumfit.DOR[[z]]$x$CQ)
  names(coef1[[z]])[c(1,4,7,10,13,14)] = c("DOR", "lnDOR", "tau^2", "tau", "I^2", "Q")
  
  fit.reitsma <- reitsma(mtdat, correction=0.5)
  fit.reitsmas[[z]] = fit.reitsma
  
  sumfit.reits[[z]] = summary(fit.reitsma)
  s = SummaryPts(fit.reitsma)
  coef2[[z]] = c(sumfit.reits[[z]]$coefficients[3,c(1,5,6)], 1-sumfit.reits[[z]]$coefficients[4,c(1,6,5)], summary(s)[4,],
                 summary(s)[1,], summary(s)[2,], sumfit.reits[[z]]$AUC$AUC)
  names(coef2[[z]])[c(1,4,7,8,11,12,15,16,19)] = c("Sensitivity", "Specificity", "Mean DOR", "Median DOR",
                                              "Mean Pos.LR", "Median Pos.LR", 
                                              "Mean Neg.LR", "Median Neg.LR", "AUC")
  names(coef2[[z]])[c(2,5,9,13,17)] = "2.5 %"
  names(coef2[[z]])[c(3,6,10,14,18)] = "97.5 %"
  
  coef[[z]] = c("Bivariate model, Reitsma 2005", round(coef2[[z]], 3), "Univariate model, Glas 2003", round(coef1[[z]], 3))
  print(paste("Protein", z, "is done."))
}

dorf = as.data.frame(do.call(rbind, coef))
dorf$Gene_name = rownames(dorf)
colnames(dorf)[c(1,21)] = "Method"
dorf = dorf[,c(38,1:37)]

colnames(dorf)[c(4:5, 7:8, 11:12, 15:16, 19:20, 24:25, 27:28, 30:31, 33:34)] =
  c("lower.Sensitivity", "upper.Sensitivity", "lower.Specificity", "upper.Specificity",
    "lower.Mean.DOR", "upper.Mean.DOR", "lower.Pos.LR", "upper.Pos.LR", "lower.Neg.LR",
    "upper.Neg.LR", "lower.DOR", "upper.DOR", "lower.lnDOR", "upper.lnDOR", 
    "lower.tau.2", "upper.tau.2", "lower.tau", "upper.tau")

dorf[,-c(1,2,22)] = apply(dorf[,-c(1,2,22)], 2, as.numeric)

```

Plots analysing the performance metrics of the SROC curves
```{r SROC curves summary, include=FALSE, echo=FALSE}
srocsig = read.csv("SROC_and_DOR_significant_proteins_2022-06-08.csv")
sroc = read.csv("SROC_and_DOR_all_proteins_2022-06-08.csv")

library(ggpubr)
library(ggpmisc)
sroc$Diff = abs(sroc$Sensitivity-sroc$Specificity)
p1 = ggplot(sroc, aes(Specificity, Sensitivity)) +
  geom_point() +
  stat_smooth(method="glm", color="red") +
  stat_cor(label.y = 0.5, method="spearman") +
  stat_poly_eq(label.y=0.25, aes(label = paste(after_stat(rr.label), sep = "*\", \"*"))) +
  scale_x_continuous(breaks=seq(0,1,0.1)) +
  scale_y_continuous(breaks=seq(0,1,0.1)) +
  theme_classic()
p1

p2 = ggplot(sroc, aes(k.study, Diff)) +
  geom_point() +
  theme_classic() +
  stat_smooth(method="glm", color="red") +
  stat_cor(label.x=2, label.y = 0.5, method="spearman") +
  stat_poly_eq(label.x=3, label.y=0.8, aes(label = paste(after_stat(rr.label), sep = "*\", \"*"))) +
  scale_x_continuous("N of studies", breaks=seq(0,12,1)) +
  scale_y_continuous("Abs. difference between Se and Sp", breaks=seq(0,1,0.1))
p2

p3 = ggplot(sroc, aes(k.study, AUC)) +
  geom_point() +
  theme_classic() +
  stat_smooth(method="glm", color="red") +
  stat_cor(label.x=8, label.y = 0.5, method="spearman") +
  stat_poly_eq(label.x=5, label.y=0.8, aes(label = paste(after_stat(rr.label), sep = "*\", \"*"))) +
  scale_x_continuous("N of studies", breaks=seq(0,12,1)) +
  scale_y_continuous("AUC", breaks=seq(0,1,0.1))
p3

p4 = ggplot(sroc, aes(k.study, I.2)) +
  geom_point() +
  theme_classic() +
  stat_smooth(method="glm", color="red") +
  stat_cor(label.x=8, label.y = 0.5, method="spearman") +
  stat_poly_eq(label.x=5, label.y=0.8, aes(label = paste(after_stat(rr.label), sep = "*\", \"*"))) +
  scale_x_continuous("N of studies", breaks=seq(0,12,1)) +
  scale_y_continuous("I2 heterogeneity", breaks=seq(0,100,10))
p4

p = ggarrange(p1, p2, p3, p4, nrow=2, ncol=2)
ggexport(p, filename = "sroc_analysis_parameters.pdf", dpi=600, width=7, height=5.5)

sroc = sroc[which(sroc$k.study>2),]

sroc$cat = "Lower Se and Lower Sp"
sroc$cat[which(sroc$AUC>=0.9)] = "90%+"
sroc$cat[which(sroc$AUC>=0.8 & sroc$AUC<0.9)] = "80-89%"
sroc$cat[which(sroc$AUC>=0.7 & sroc$AUC<0.8)] = "70-79%"
sroc$cat[which(sroc$AUC<0.7)] = "<69%"
sroc$cat[which(!sroc$Gene_name %in% srocsig$Gene_name)] = "Not significant"
table(sroc$cat)
sroc$cat = as.factor(sroc$cat)
sroc$cat = ordered(sroc$cat, levels=levels(sroc$cat)[c(4,1:3)])

sroc2 = sroc[sroc$cat=="80-89%",]
p = ggplot(sroc, aes(1-Specificity, Sensitivity, col=cat)) +
  geom_point() +
  scale_x_continuous("FPR (1-Specificity)", limits=c(0,1), breaks=seq(0,1,0.1)) +
  scale_y_continuous("Sensitivity", limits=c(0,1), breaks=seq(0,1,0.1)) +
  theme_classic() +
  geom_segment(x = 0, y = 0, xend = 1, yend = 1, col="black", linetype="dashed", size=1) +
  scale_color_manual(name="SROC AUC", values=c(alpha("lightgray", 0.4), "#E69F00", "#009E73", "#D55E00")) +
  theme(legend.position = c(0.8, 0.15)) +
  geom_text_repel(data=sroc2, label=sroc2$Gene_name, max.overlaps = 30)
p

ggsave("Meta_analysis_SROC_curves_significant.pdf", p, dpi=600, width=7, height=5.5)

```

Plot the SROC curves for the top 10 proteins
```{r top10_proteins, include=FALSE, echo=FALSE}
# Load the FN, TP, FP, TN for each protein in each study
mtdats = read_rds("mtdats.Rdata")

#### Choose sproteins of interest
sel = sroc2$Gene_name[which(sroc2$k.study==11 & sroc2$AUC>0.835)]
s = list()
srs = list()
rdf = list()
sspr = list()
auc = list()
sf = list()

for (z in sel) {
  mtdatr = mtdats[[z]]
  #### Forest plots
  nam = studies[as.numeric(rownames(mtdatr))]
  fit.reitsma = reitsma(mtdatr)
  
  #### Extract data for the SROCs
  s[[z]] = summary(fit.reitsma)
  sr = sroc(fit.reitsma) ## To get the SROC curve data
  msr = mcsroc(fit.reitsma)
  re = ROCellipse(fit.reitsma, level=0.95)
  rdf[[z]] = as.data.frame(re$ROCellipse)
  
  srs[[z]] = as.data.frame(sr)
  sspr[[z]] = as.data.frame(s[[z]]$coefficients[3:4,c(1,5:6)])
  
  a = AUC(fit.reitsma)
  auc[[z]] = a$AUC
  
  sfit = SummaryPts(fit.reitsma)
  sf[[z]] = summary(sfit)
}

srsz = as.data.frame(do.call(rbind, srs))
srsz$protein = gsub("[.].+", "", rownames(srsz))
ssprz = as.data.frame(do.call(rbind, sspr))
ssprz$protein = gsub("[.].+", "", rownames(ssprz))

scof = list()
for (i in names(s)) {
  scof[[i]] = s[[i]]$coefficients[c(4,3)]
}
sz = as.data.frame(do.call(rbind, scof))
sz$protein = gsub("[.].+", "", rownames(sz))
rdfz = as.data.frame(do.call(rbind, rdf))
rdfz$protein = gsub("[.].+", "", rownames(rdfz))

colors = brewer.pal(n = 10, name = "Paired")

##### Plot the summary sensitivity and specificity per protein
p = ggplot(srsz, aes(fpr, V2, col=protein)) +
  geom_mark_hull(data=rdfz, aes(V1, V2, col=protein), expand=0, linetype="dotted", size=0.75) +
  geom_line(size=1) +
  scale_x_continuous("FPR (1-Specificity)", limits=c(0,1), breaks=seq(0,1,0.1)) +
  scale_y_continuous("Sensitivity", limits=c(0,1), breaks=seq(0,1,0.1)) +
  theme_classic() +
  geom_segment(x = 0, y = 0, xend = 1, yend = 1, col="black", linetype="dashed", size=1) +
  geom_point(data=sz, aes(x=V1, y=V2, col=protein, fill=protein), shape=23, size=3) +
  theme(legend.position = c(0.8, 0.3)) +
  scale_color_manual("Protein", values=colors) +
  scale_fill_manual("Protein", values=colors)
p

ggsave("Meta_analysis_SROC_top_10.pdf", p, dpi=600, width=7, height=5.5)

```
