---
title: "covid_hirief"
author: "Haris Babačić"
date: "08/12/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

Run this chunk to add libraries and datasets
```{r setup, include=FALSE}
pack <- c("tidyverse", "ggplot2", "matrixStats", "VennDiagram", "ggrepel", "dplyr", "factoextra", "FactoMineR", "ggfortify", "cluster", "gplots", "RColorBrewer", "ComplexHeatmap", "readxl", "ggpubr", "msigdbr", "grid", "limma", "circlize",
          "org.Hs.eg.db", "clusterProfiler", "enrichplot", "WGCNA")
lapply(pack, require, character.only = TRUE)
source("https://gist.githubusercontent.com/jokergoo/bfb115200df256eeacb7af302d4e508e/raw/14f315c7418f3458d932ad749850fd515dec413b/word_cloud_grob.R")

# Change the working directory to correspond to the location of the folder
setwd("code_resubmission/data")

data = read.delim("genes_table_serum.txt", header=T) # To analyse the data in a gene-centric manner
# If you want to get the SARS-CoV-2 proteins, you need to use the protein-centric table below
# data = read.delim("proteins_table_serum", header=T) # To analyse the data in a protein-centric manner
colnames(data)
# Clinical data are not available due to legal constraints

# Select the variables of interest
sel = c(15:46)
### To run the protein tables search with the SARS-CoV-2 peptides deselect the lines with # below
#grep("_tmt16plex_", colnames(data))  
df = as.matrix(data[,sel])
rownames(df) = data$Gene.Name
colnames(df) = gsub("_set1_.+", "", colnames(df))
colnames(df) = gsub("_set2_.+", "", colnames(df))
colnames(df) = gsub("X__POOL_", "", colnames(df))

# Remove TMT pools
rem = grep("IS", colnames(df))
df = df[, -rem]

# Add the COVID to the label IDs, to fit the clinical dataset
colnames(df)[-grep("HC", colnames(df))] = paste("COVID", colnames(df)[-grep("HC", colnames(df))], sep="")
# Save the dataset
write.csv(df, "data_matrix.csv")

# Load the Filbin and SOMAscan data
pea = read_xlsx("filbin_cell_reports_medicine_2021_table_s3.xlsx", sheet=3)
length(intersect(pea$Assay, data$Gene.Name))
soma = read.csv("sullivan_cell_reports_2021_somascan_summary.csv")
soma$gene = gsub("[.].+", "", soma$Analyte)
length(intersect(soma$gene, data$Gene.Name))

# Load the Human Protein Atlas (HPA) datasets for protein annotation
cetype = read_tsv("hpa/cell_type_category_rna_any_Cell.tsv")
grtype = read_tsv("hpa/cell_type_category_rna_any_Group.tsv")
enh = read_tsv("hpa/cell_type_category_rna_any_Cell_enhanced.tsv")
tisscat = read_tsv("hpa/tissue_category_rna_any_Tissue.tsv")

```

```{r create_functions, include=FALSE, echo=FALSE}
```


Comparison COVID-19 to PCR-negative controls
```{r differential_expression, include=FALSE, echo=FALSE}
# Change the directory to the results folder
setwd("../results")

# Check the distribution of the data
hist(df, breaks=100)
boxplot(df)

# Calculate the missing values in percentages
miss = apply(df, 1, function(x) length(which(is.na(x)))/length(x)*100)
names(miss) = data$Gene.Name
table(miss<50)

# Select all the proteins with 50% observations in the data
sel = names(miss[which(miss<50)])
dfaf = df[sel,]

##### Perform unpaired t test comparing COVID-19 patients and healthy controls
# Select the COVID-19 cases
cov = grep("COVID", colnames(dfaf))
# Select the healthy controls
hc = grep("HC", colnames(dfaf))

# Perform unpaired two-sided t test
t_res = apply(dfaf, 1, function(x) t.test(x[cov], x[hc]))
tres = data.frame(matrix(unlist(t_res), nrow=length(t_res), byrow=T))
tres[,1:9] = apply(tres[,1:9], 2, as.numeric)
colnames(tres) <- c("statistic", "df", "p.value", "conf_int_lower", "conf_int_upper", "estimate_mean_x", "estimate_mean_y", "null.value", "stderr", "side", "method", "data.name")

tres$gene = names(t_res)
tres$protein = data$Protein.ID.s.[match(tres$gene, data$Gene.Name)]
tres$adj.p = p.adjust(tres$p.value, method="fdr")
tres$fc = rowMeans(dfaf[,cov], na.rm=T)-rowMeans(dfaf[,hc], na.rm=T)

# Check the distribution of p and q values 
hist(tres$p.value, breaks=100)
hist(tres$adj.p, breaks=100)

# Subset the differentially altered proteins (DAPs) at 5% FDR and save the results
diffall = tres[which(tres$adj.p<0.05),]
write.csv(tres, "covid_vs_hc_all_res.csv")
write.csv(diffall, "covid_vs_hc_difall.csv")

```

Volcano plot for the t test results
```{r volcano_t_test, include=FALSE, echo=FALSE}
tres$alteration = "Up in Covid-19"
tres$alteration[which(tres$fc<0)] = "Down in Covid-19"

tres$shade = abs(tres$fc)*-log10(tres$adj.p)
tres$shadcat = cut(tres$shade, breaks=c(0,0.25, 0.5, 0.75, 1, 1.5, 2.5, 7))
levels(tres$shadcat) = c(1:7)
table(is.na(tres$shadcat))
table(tres$shadcat)

y <- tres[which(tres$shade>2.5 & tres$fc>0),]
r <- tres[which(tres$shade>2.5 & tres$fc<0),]

cov = ggplot(tres, aes(x=fc, y=-log10(adj.p), col=alteration)) + 
    geom_point(aes(alpha=as.numeric(shadcat))) + 
    theme_classic() + 
    ylab(expression(paste("-log10 ", italic("q"), " values"))) + 
    xlab("log2 fold-change") + 
    scale_alpha(range=c(0.1,1), guide="none") +
    geom_hline(yintercept=-log10(0.05), col="red", linetype="dotdash", linewidth=1) + 
    scale_color_manual("Alteration", values=c("#0072B2", "#D55E00")) +
    theme(legend.position = c(0.9, 0.1)) +
    geom_text_repel(data=y, aes(x=fc, y=-log10(adj.p), label=gene), col="#D55E00", vjust=0.5, hjust=0.1) + 
    geom_text_repel(data=r, aes(x=fc, y=-log10(adj.p), label=gene), col="#0072B2", vjust=0.5, hjust=-0.1) + 
    scale_x_continuous("", breaks=seq(round(min(tres$fc)),round(max(tres$fc)), 0.5)) +
    scale_y_continuous("", breaks=seq(round(min(-log10(tres$adj.p), digits=1)), 
                                  round(max(-log10(tres$adj.p)), digits=1), 1))
cov

ggsave("cov_volc.png", cov, width=6.25, height=5, units="in", dpi=600)

quartz(type = 'pdf', file = 'cov_volc.pdf', width=6.25, height=5)
cov
dev.off()

```

Adjust the comparison COVID-19 vs HC for clinical variables that have enough events
- This chunk will give errors because the clinical dataset is not available due to legal constraints.
- We provide the code used for the analysis
```{r limma_model, include=FALSE, echo=FALSE}
### Limma
x = as.data.frame(t(dfaf))
rownames(x) = gsub("X", "", rownames(x))
x$type = gsub("[0-9]+", "", rownames(x))
x$type = as.factor(x$type)

clim = cd[which(cd$`Patient ID` %in% rownames(x)),]
sel = match(rownames(x), clim$`Patient ID`)
clim$type = gsub("[0-9]+", "", clim$`Patient ID`)

var = clim[sel, colnames(clim)[c(2:4, 6:7, 44)]]
m = as.matrix(x[,-ncol(x)])

# Impute missing values
apply(var, 2, function(x) length(which(is.na(x)))/length(x))
var$Hypertension[which(is.na(var$Hypertension))] = 0
var$Hypertension[which(var$Hypertension=="NA")] = 0
var$`Type II diabetes`[which(is.na(var$`Type II diabetes`))] = 0
var$`Type II diabetes`[which(var$`Type II diabetes`=="NA")] = 0
var$Astma[which(is.na(var$Astma))] = 0
var$Astma[which(var$Astma=="NA")] = 0
colnames(var)[4] = "Diabetes"
str(var)
var[,c(3:5)] = apply(var[,c(3:5)], 2, as.factor)

### Limma
#design = model.matrix(~ 0+type + Age + Sex + Albumin + nlr, var)
#colnames(var)[c(10,12)] = c("SOFA_resp", "Invasive_ventilation")
design = model.matrix(~ 0 + type + Age + Sex + Hypertension + Diabetes, var)
colnames(design)[c(1:2)] = gsub("type", "", colnames(design)[c(1:2)])

fit1 = lmFit(t(m), design)

x = c("COVID-HC")
contrast =  makeContrasts(contrasts=x,levels=design)
fit2 <- contrasts.fit(fit1, contrasts = contrast)
fit3 = eBayes(fit2)

### Extract the results from limma
results_limma = topTable(fit3, number=ncol(m))
results_limma$gene = rownames(results_limma)
res = results_limma[which(results_limma$adj.P.Val<0.05),]

write.csv(results_limma, "results_limma_all_adjusted_age_sex_hypertension_asthma_diabetes.csv")
write.csv(res, "results_limma_all_adjusted_age_sex_hypertension_asthma_diabetes_significant.csv")

```

Overlap of proteins identified by HiRIEF and affinity-based proteomics
```{r protein_overlap_methods, include=FALSE, echo=FALSE}
##### Average PSMs and MS1 precursor area #####
psm = data[,grep("Amount.PSMs", colnames(data))]
data$psm_average = apply(psm, 1, function(x) mean(x, na.rm=T))
data$psm_median = apply(psm, 1, function(x) median(x, na.rm=T))

pa = data[,grep("MS1.precursor.area", colnames(data))]
data$pa_average = apply(pa, 1, function(x) mean(x, na.rm=T))
data$pa_median = apply(pa, 1, function(x) median(x, na.rm=T))

mydata_2 = data
mydata_2 = mydata_2[order(mydata_2$pa_median, decreasing = T),]
mydata_2$number = c(1:nrow(mydata_2))

##### Overlap between HiRIEF, Olink, and SOMAscan #####
library(ggvenn)

x <- list(
  "HiRIEF LC-MS/MS" = mydata_2$Gene.Name, 
  "SomaScan Aptamers" = soma$gene,
  "Olink Explore PEA" = pea$Assay
)

ven = ggvenn(x, 
             fill_color = c("#E69F00", "#D55E00", "#56B4E9", "#009E73"),
             stroke_size = 0.25, set_name_size = 4, show_percentage = F)  +
  theme(plot.margin = unit(c(0,0,0,0), "cm"))
ven

ggsave("ven_gsaps.pdf", ven, dpi=300, width=4, height=3)
ggsave("ven_gsaps.jpeg", ven, dpi=600, width=5, height=3.5)

##### MS1 precursor area plot #####
mydata_2$Method = "Only HiRIEF"
mydata_2$Method[which(mydata_2$Gene.Name %in% pea$Assay)] = "Olink PEA and HiRIEF"
mydata_2$Method[which(mydata_2$Gene.Name %in% soma$gene)] = "SomaScan and HiRIEF"
mydata_2$Method[which(mydata_2$Gene.Name %in% intersect(soma$gene, pea$Assay))] = "All three"
table(mydata_2$Method)

selected = which(mydata_2$Method=="Only HiRIEF")
gs = mydata_2[-selected,]
table(mydata_2$Method)

chosen = which(mydata_2$Gene.Name %in% c("LBP", "CRP", "GOT1", "LDHA", "LDHB", "GPT", "CRTAC1", "VWF", "F13A", "MX1","ISG15",
                                         "GOLM1","C9", "GSN","SHBG", "CD276",  "ADA","CXCL16", "S100A8", "MMP8", "CXCL12", "SRPK2",
                                         "PDCD6",  "APMAP", "SHH","SFTB", "ORM1", "ORM2", "ADA2", "DCTN1", "F11R", "F2R","GGT1",
                                         "SAA2", "SAA4", "C1QB", "C8B","MYOZ1",  "TTR", "HSPA4", "PROCR", "H1-5", "SERPINA1"))
annot = mydata_2[chosen,]

ms1_area = ggplot(mydata_2[-selected,], aes(x=number, y=log10(pa_median))) +
  geom_point(alpha=0.5) +
  theme_classic() +
  geom_point(data=gs[-chosen,], aes(x=number, y=log10(pa_median), color=Method), size=2) +
  geom_text_repel(data=annot, aes(x=number, y=log10(pa_median), 
                               color=Method, label=Gene.Name), vjust=0.5, hjust=-0.5,
                  direction = 'y',
                  nudge_y = 0.6,
                  segment.size=0.2,
                  max.overlaps = 20) +
  annotation_logticks() +
  scale_x_continuous(name="Protein rank", breaks=seq(0,3500,500)) +
  geom_point(data=annot, aes(x=number, y=log10(pa_median), color=Method), size=3) +
  scale_y_continuous(name="MS1 precursor area (log10)", breaks=seq(5,13,1)) +
  scale_color_manual(values=c("#009E73", "#D55E00", "black", "#56B4E9")) +
  theme(legend.position = c(0.8, 0.8))
ms1_area  

ggsave("ms1_area_methods.pdf", ms1_area, dpi=300, height=5, width=7)
ggsave("ms1_area_methods.jpeg", ms1_area, dpi=600, height=5, width=7)

```

PCA and clustering (PAM and hierarchical)
```{r clustering_all_proteins, include=FALSE, echo=FALSE}
dfa = na.omit(df)
# Try clustering the serum data
all <- t(dfa)
rownames(all)[hc] = paste("HC", 1:7, sep="")
rownames(all)[-hc] = paste("COVID", 1:20, sep="")
all = apply(all, 2, as.numeric)
rownames(all) = colnames(dfa)

# Principal component analysis (PCA)
pcad <- prcomp(all, scale. = TRUE)
score.df <- as.data.frame(pcad$x)

se.pca <- PCA(all, graph=TRUE)
plot(se.pca)

# PCA
theme_set(theme_bw(base_size=12)) # set default ggplot2 theme

all <- as.data.frame(all)
all$type <- NA
all$type[hc] <- "HC"
all$type[-hc] <- "COVID"
table(all$type)

all %>% dplyr::select(-type) %>% scale() %>% prcomp() -> pca
all %>% dplyr::select(-type)  %>% kmeans(centers=2, nstart=10) -> km

pdf("no_na_pca.pdf", height=4, width=5)
autoplot(pam(all[,-which(colnames(all)=="type")], 3), data=all, 
         colour="type", frame = TRUE, frame.type = 'norm',
         shape="type", size=3) + theme_classic() +
  scale_color_manual(values=c("#D55E00", "#009E73")) +
  scale_x_continuous(breaks=seq(-0.5,0.5,0.1)) +
  scale_y_continuous(breaks=seq(-0.8,0.8,0.2)) +
  theme(legend.position = c(0.9,0.9))
dev.off()

v = summary(as.numeric(dfa))
colup = "#E69F00"
coldown = "#0072B2"
mid = "white"
col_fun = colorRamp2(c(v[1], -1, v[3], 1, v[6]), 
                     c(coldown, alpha(coldown, 0.8), mid, alpha(colup, 0.8), colup))
col_fun(seq(-3, 3))

theme_set(theme_bw(base_size=6))
jpeg("hierarchical_all.jpeg", height=6, width=10, units="in", res=600)
Heatmap(dfa, col = col_fun, clustering_distance_rows = "spearman", show_row_names = FALSE)
dev.off()

```

Make a complete heatmap with annotations for differentially altered proteins
- This chunk will give errors because the clinical dataset is not available due to legal constraints.
```{r heatmap_differential_expression, include=FALSE, echo=FALSE}

a = all[,which(colnames(all) %in% diffall$gene)]
#colnames(a) = diffall$gene[match(colnames(a), diffall$protein)] #### For protein table
a = t(a)
colnames(a) = gsub("X", "", colnames(a))

h = Heatmap(a, col = col_fun, clustering_distance_rows = "spearman")
ht = draw(h)
row_order(ht)

nam = rep("", nrow(a))
lab = rownames(a)[row_order(ht)]
nam[seq(1, nrow(a), 5)] = lab[seq(1, nrow(a), 5)]
names(nam) = lab

cd = cd[match(colnames(a), cd$`Patient ID`),]
table(cd$`Patient ID`==colnames(a))
c = c[match(colnames(a), c$ID),]
table(cd$`Patient ID`==colnames(a))
table(c$ID==colnames(a))

group = cd$`Patient ID`
group = gsub("[0-9]+", "", group)
group[which(group=="HC")] = "Healthy"
drug = as.factor(c$drug)
levels(drug) = c(NA, "Anakinra", "Chloroquine", NA)
bron = as.factor(c$bronchodilation)
levels(bron) = c("No", "Yes", "No")
fr = as.factor(c$fragmin)
levels(fr) = c("No", "Yes, 5,000 IU", "Yes, 7,500 IU", "No")
fev = as.factor(c$feber)
levels(fev) = c("Yes", "Yes", "No")
vir = as.factor(c$Viremi)
levels(vir) = c("No", "Yes")
died = as.factor(cd$`Fatal outcome`)
levels(died) = c("No", "Yes", "No")
hta = as.factor(cd$Hypertension)
levels(hta) = c("No", "Yes", "No")
cvd = as.factor(cd$`Cardiovascular disease`)
levels(cvd) = c("No", "Yes", "No")
astma = as.factor(cd$Astma)
levels(astma) = c("No", "Yes", "No")
diab = as.factor(cd$`Type II diabetes`)
levels(diab) = c("No", "Yes", "No")
sev = as.factor(cd$`Disease severity based on O2 max`)
levels(sev) = c("Critically ill", NA, "Severe")

v = summary(as.numeric(a))
colup = "#E69F00"
coldown = "#0072B2"
mid = "white"
col_fun = colorRamp2(c(v[1], -2, v[3], 2, v[6]), 
                     c(coldown, alpha(coldown, 0.5), mid, alpha(colup, 0.5), colup))

numvar = colnames(cd)[c(3, 9, 11, 14:43)]
sum = list()
colfun = list()
for (i in numvar[-19]) {
  sum[[i]] = summary(as.numeric(unlist(cd[,i])))
  s = as.numeric(sum[[i]])
  colfun[[i]] = colorRamp2(c(s[1], s[6]), c("white", "#D55E00"))
}

ha = HeatmapAnnotation(Group = group,
                       Sex = cd$Sex,
                       Age = cd$Age,
                       "Days to sampling" = as.numeric(cd$`Symptom onset to sampling (days)`),
                       Severity = sev,
                       "Days of hospitalisation" = as.numeric(cd$`Hospitalization duration (days)`),
                       Died = died,
                       Hypertension = hta,
                       "Type II diabetes" = diab,
                       Asthma = astma,
                       CVD = cvd,
                       Fever = fev,
                       Viremia = vir,
                       CRP = cd$CRP,
                       "D-dimers" = cd$`D-dimer`,
                       IL6 = as.numeric(cd$`IL-6 (pg/mL)`),
                       NLR = cd$NLR,
                       "Received drug" = drug,
                       "Received Deltaparin" = fr,
                       "O2 max liters/minute" = as.numeric(cd$`O2 max liters/minute`),
                       col=list(Group=c("COVID"="#D55E00", "Healthy"="#009E73"),
                                Sex=c("F"="#CC79A7", "M"="#0072B2"),
                                Severity=c("Critically ill"="tomato2", "Severe"="royalblue3"),
                                Died=c("Yes"="black", "No"="white"),
                                Hypertension = c("Yes"="darkgreen", "No"="white"),
                                "Type II diabetes" = c("Yes"="goldenrod3", "No"="white"),
                                Asthma = c("Yes"="chartreuse3", "No"="white"),
                                CVD = c("Yes"="coral", "No"="white"),
                                Fever = c("Yes"="seagreen", "No"="white"),
                                Viremia = c("Yes"="indianred3", "No"="white"),
                                "Received drug" = c("Anakinra"="yellow3", "Chloroquine"="lightskyblue3"),
                                "Received Deltaparin" = c("No"="white", "Yes, 5,000 IU"="darkgreen", 
                                                          "Yes, 7,500 IU"="goldenrod3"),
                                Age=colfun[["Age"]],
                                "Days of hospitalisation"=colfun[["Hospitalization duration (days)"]],
                                "O2 max liters/minute"=colfun[["O2 max liters/minute"]],
                                CRP=colfun[["CRP"]],
                                "D-dimers"=colfun[["D-dimer"]],
                                IL6=colfun[["IL-6 (pg/mL)"]],
                                NLR=colfun[["NLR"]]
                                ), na_col="white")

sel = which(rownames(a) %in% mydata_2$Gene.Name[chosen])
ra = rowAnnotation(foo = anno_mark(at = sel, labels = rownames(a)[sel]), annotation_name_rot = 45)

#a = a[sel,]

ha2 = HeatmapAnnotation(Procalcitonin = cd$`Pro-calcitonin`,
                       TPK = cd$TPK,
                       Ferritin = cd$Ferritin,
                       AST = cd$ASAT,
                       ALT = cd$ALAT,
                       LDH = cd$LDH,
                       "Troponin T" = cd$`troponin T`,
                       Hb = cd$Hb,
                       Neutrophils = cd$Neutrophils,
                       Lymphocytes = cd$Lymphocytes,
                       #Basophils = cd$Basophils,
                       Eosinophils = cd$Eosinophils,
                       Monocytes = cd$Monocytes,
                       BAFF = cd$`BAFF (pg/mL)`,
                       "Antibody secreting cell numbers/uL" = cd$`Antibody secreting cell numbers/uL blood`,
                       "% of antibody secreting B cells" = cd$`% of antibody secreting cells of all B cells`,
                       "Neutralizing antibodies (titers)" = cd$`Neutralizing antibodies (titers)`,
                       "% of activated CD4" = cd$`% of activated CD4`,
                       "% of activated CD8" = cd$`% of activated CD8`,
                       "IgM against N" = cd$`IgM against N (OD450nm) EDI`,
                       "IgA against S1" = cd$`IgA against S1 (OD ratio) Euroimmun`,
                       "IgG against S1" = cd$`IgG against S1 (OD ratio) Euroimmun`,
                       "IgG against whole SARS-CoV-2 (titre)" = cd$`IgG against whole SARS-CoV-2 (titer) immunofluorescence`,
                       na_col="white")


hm = Heatmap(a,
        clustering_distance_rows="spearman", 
        name="log2",
        column_names_side = "top",
        show_row_names = F, 
        show_row_dend = F,
        column_dend_side = "bottom", 
        show_column_names = F, 
        column_split = 4,
        row_split = 3,
        border = T,
        col = col_fun,
        #top_annotation = ha # The heatmap can be obtained without the clinical information
        )

ht = draw(hm)
genes_down = rownames(a)[row_order(ht)[[1]]]
genes_up_1 = rownames(a)[row_order(ht)[[2]]]
genes_up_2 = rownames(a)[row_order(ht)[[3]]]

theme_set(theme_bw(base_size=6))
pdf("heatmap_differentially_expressed_in_covid_.pdf", height=10, width=16)
draw(hm, heatmap_legend_side = "right", 
    annotation_legend_side = "right",
    padding = unit(c(0.5, 2.5, 0.5, 3.5), "cm"))
dev.off()

```

Heatmap annotation based on different databases
- This chunk will give errors because the clinical dataset is not available due to legal constraints.
```{r go_heatmap_annotation, include=FALSE, echo=FALSE}

reactome <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "REACTOME") %>% 
  dplyr::select(gs_name, entrez_gene, gene_symbol)
head(reactome)

#### Remove the REACTOME prefix - redundant
reactome$gs_name = gsub("REACTOME_", "", reactome$gs_name)
reactome$gs_name = gsub("_", " ", reactome$gs_name)

prot = list()
prgo = list()
for(i in 1:3) {
  prot[[i]] = rownames(a)[row_order(ht)[[i]]]
  prgo[[i]] = reactome$gs_name[which(reactome$gene_symbol %in% prot[[i]])]
}

keywords = list()
for (i in 1:3) {
  k = as.data.frame(table(prgo[[i]]))
  colnames(k) = c("word", "freq") 
  k = k[order(k$freq, decreasing = T),] 
  keywords[[i]] = k[1:5,]
}
names(keywords) = c(1:3)

cl = rep(1, nrow(a))
cl[row_order(ht)[[2]]] = 2
cl[row_order(ht)[[3]]] = 3
table(cl)

cl = as.factor(cl)

ht2 = Heatmap(a, col = col_fun,
             name = "log2",
             show_row_names = FALSE, show_column_names = FALSE,
             show_row_dend = F, show_column_dend = T,
             row_split = cl, column_split = 4, 
             border = "#404040", row_title = NULL, column_title = NULL,
             #top_annotation = ha, The heatmap can be obtained without clinical information
             row_gap = unit(0, "mm"), column_gap = unit(0, "mm"))

align_to = split(seq_len(nrow(a)), cl)
align_to = align_to[names(align_to) != "0"]
align_to = align_to[names(align_to) %in% names(keywords)]
align_to

fontsize_range = c(4, 16)
gbl = lapply(names(align_to), function(nm) {
  kw = keywords[[nm]][, 1]
  freq = keywords[[nm]][, 2]
  fontsize = scale_fontsize(freq, rg = c(1, max(10, freq)), fs = fontsize_range)
  
  word_cloud_grob(text = kw, fontsize = fontsize)
})
names(gbl) = names(align_to)
gbl

margin = unit(8, "pt")
gbl_h = lapply(gbl, function(x) convertHeight(grobHeight(x), "cm") + margin)
gbl_h = do.call(unit.c, gbl_h)

gbl_w = lapply(gbl, function(x) convertWidth(grobWidth(x), "cm"))
gbl_w = do.call(unit.c, gbl_w)
gbl_w = max(gbl_w) + margin

panel_fun = function(index, nm) {
  # background
  grid.rect(gp = gpar(fill = "white", col = NA))
  # border
  grid.lines(c(0, 1, 1, 0), c(0, 0, 1, 1), gp = gpar(col = "#AAAAAA"), 
             default.units = "npc")
  gb = gbl[[nm]]
  # a viewport within the margins
  pushViewport(viewport(x = margin/2, y = margin/2, 
                        width = grobWidth(gb), height = grobHeight(gb),
                        just = c("left", "bottom")))
  grid.draw(gb)
  popViewport()
}

ht3 = ht2 + rowAnnotation(keywords = anno_link(align_to = align_to, 
                                             which = "row", panel_fun = panel_fun, 
                                             size = gbl_h, gap = unit(2, "mm"), 
                                             width = gbl_w + unit(5, "mm"), # 5mm for the link
                                             link_gp = gpar(fill = "white", col = "#AAAAAA"), 
                                             internal_line = FALSE)) # you can set it to TRUE to see what happens

#theme_set(theme_bw(base_size=8))
pdf("heatmap_differentially_expressed_in_covid_annotated_reactome.pdf", height=8, width=16)
draw(ht3, ht_gap = unit(3, "pt"))
dev.off()

```

Code for comparison to PEA and SOMAscan findings - exemplified with Olink
```{r olink_comparison, include=FALSE}
# Filbin study
### The comments refer to changes required for the comparison to limma results
# Change tres to results_limma$gene if you want to compare the agreement between PEA and limma results.
ov = intersect(pea$Assay, tres$gene) 
res = tres # Change to results_limma 
res = res[which(res$adj.p<0.05),] # Change here based on the significance wanted

res$olink = pea$`estimate (difference in expression given as COVID-positive minus COVID-negative)`[match(res$gene, pea$Assay)]
pea$adj.pval = p.adjust(pea$p.value, method="fdr")
# Change here based on significance
oreSS = pea[which(pea$adj.pval<0.05),]
res$olsign = oreSS$`estimate (difference in expression given as COVID-positive minus COVID-negative)`[match(res$gene, oreSS$Assay)]

res$olcat = NA # Change to results_limma$logFC for fold change
res$olcat[which(res$fc>0 & res$olsign>0)] = paste('\U2191', "HiRIEF-MS,", '\U2191', "Olink-PEA")
res$olcat[which(res$fc>0 & res$olsign<0)] = paste('\U2191', "HiRIEF-MS,", '\U2193', "Olink-PEA")
res$olcat[which(res$fc<0 & res$olsign<0)] = paste('\U2193', "HiRIEF-MS,", '\U2193', "Olink-PEA")
res$olcat[which(res$fc<0 & res$olsign>0)] = paste('\U2193', "HiRIEF-MS,", '\U2191', "Olink-PEA")
t = table(res$olcat)
a = sum(t[c(1,4)])/sum(t)*100

cv = cor(res$fc, res$olsign, use="pairwise.complete.obs", method="spearman")
cv

# Change this to adj.P.Val for limma
sel = which(res$adj.p<0.05)

p = ggplot(res[sel,], aes(fc, olsign, col=olcat)) +
  geom_point() +
  geom_text_repel(label=res[sel,]$gene) + 
  theme_classic() +
  geom_vline(xintercept = 0, col="darkgray", linetype="dashed", size=0.5) +
  geom_hline(yintercept = 0, col="darkgray", linetype="dashed", size=0.5) +
  scale_x_continuous("HiRIEF LC-MS/MS, log2-FC COVID-19:Healthy", breaks=seq(-2,2,0.25)) +
  scale_y_continuous("Olink Explore PEA, log2-fc COVID-19:Controls", breaks=seq(-2,2,0.25)) +
  annotate("text", x=-1.5, y=1.5, hjust=0, col="black",
           label=paste("Agreement: ", round(a, digits=3), "%", sep="")) +
  annotate("text", x=-1.5, y=1.3, hjust=0, col="black",
           label=paste("r = ", round(cv, digits=3), sep="")) +
  scale_color_manual(name="change", values=c("#D55E00", "#E69F00", "#CC79A7", "#009E73")) +
  theme(legend.position = c(0.8, 0.15))
p

quartz(type = 'pdf', file = 'Filbin_Olink_adj_p_values_vs_HiRIEF_adj_pval.pdf', width=7, height=5)
p
dev.off()
ggsave("Filbin_Olink_adj_p_values_vs_HiRIEF_adj_pval.jpeg", p, dpi=600, width=8, height=8)

```

Annotate which proteins were identified by the other two methods, their significance, and tissue-enrichment
```{r annotate_significance_and_tissue_origin, include=FALSE, echo=FALSE}
##### Annotate which one is significant where #####
tressign = tres[which(tres$adj.p<0.05),]
tisscat$spec_tissue = gsub(":.+", "", tisscat$`RNA tissue specific nTPM`)
tressign$tissspec = tisscat$spec_tissue[match(tressign$gene, tisscat$Gene)]
table(tressign$tissspec)

tressign$PEA = "Not significant"
table(pea$adj.pval<0.05, pea$`Adjusted_pval (after MHT)`<0.05)
tressign$PEA[which(tressign$gene %in% pea$Assay[which(pea$adj.pval<0.05)])] = "Significant"
tressign$PEA[which(!tressign$gene %in% pea$Assay)] = "Not targeted by PEA"
t = table(tressign$PEA)
t[3]/(t[1]+t[3])

tressign$SOMAscan = "Not significant"
tressign$SOMAscan[which(tressign$gene %in% soma$gene[which(soma$adj.pvalue<0.05)])] = "Significant"
tressign$SOMAscan[which(!tressign$gene %in% soma$gene)] = "Not targeted by aptamers"
t2 = table(tressign$SOMAscan)
t2[3]/(t2[1]+t2[3])

sel = c(13,14,16,15,17,19:21)
colnames(tressign)[sel]

tres_anot = tressign[,sel]

colnames(tres_anot) = c("Gene name", "Protein IDs", "log2-FC", "FDR q value", "Alteration", "HPA tisue-enrichment", "PEA quant.", "SOMAscan quant.")
write.csv(tres_anot, "annotation_significant_proteins.csv")

```

Code for correlation to clinical assays of selected proteins
- This chunk will give errors because the clinical dataset is not available due to legal constraints.
```{r correlation_chemistry, include=FALSE, echo=FALSE}
colnames(df) = gsub("X", "", colnames(df))
table(colnames(df) == cd$`Patient ID`)

#### CRP
sel = "CRP"

x = df[sel,]
x = x[grep("COVID", names(x))] # If you want only Covid patients
y = cd$CRP[match(names(x), cd$`Patient ID`)]

cor(x, log2(y), method="spearman")
df1 = as.data.frame(cbind(as.numeric(x), as.numeric(y)))

formula = y~x

corr_crp = ggplot(df1, aes(V1, log2(V2))) +
  geom_point() +
  theme_classic() +
  stat_cor(method="spearman", label.x=-0.5, label.y=7.7, hjust=0.5) +
  stat_smooth(method = "lm", formula=formula, col="darkgreen") +
  stat_regline_equation(aes(label =  paste(..adj.rr.label..)), formula=formula, label.x=-0.5, label.y=7.4, hjust=1) +
  scale_x_continuous(name = "CRP (log2) HiRIEF LC-MS/MS", breaks=seq(-1,1,0.25)) + 
  scale_y_continuous(name="CRP (log2) immunoassay", breaks=seq(4.5,8,0.5))
corr_crp

ggsave("corr_crp.pdf", corr_crp, width=4, height=3)

```

Gene set enrichment analyses for the differentially-altered proteins
```{r enrichment, include=FALSE, echo=FALSE}
# Create a vector of proteins' log2-FC
geneList <- tres$fc

## Assign protein names to the numeric vectors
names(geneList) <- tres$gene

## Sort the proteins based on their log2-FC
geneList <- sort(geneList, decreasing = TRUE)
length(geneList)

# Extract the proteins' ENTREZ IDs
gene.df <- bitr(names(geneList), fromType = "SYMBOL",
        toType = c("ENSEMBL", "ENTREZID"),
        OrgDb = org.Hs.eg.db)
head(gene.df)

# Replace the gene names with their ENTREZ IDs
geneList2 <- geneList
names(geneList2) <- gene.df$ENTREZID[match(names(geneList2), gene.df$SYMBOL)]
names(geneList2)

# Remove duplicates that can give uncertainty in the ranking
table(duplicated(names(geneList2)))
geneList3 <- geneList2[-which(duplicated(names(geneList2))==TRUE)]

# Perform GSEA analysis
# Specify the MSigDb dataset you need. Here you can specify the entire dataset with all genes
m_df <- msigdbr(species = "Homo sapiens") %>% 
  dplyr::select(gs_name, entrez_gene, gene_symbol)
head(m_df, 2) %>% as.data.frame

# Selecting the chemical-genetic perturbations database
# You can change here the category and subcategory of interest for the different analyses, for e.g. C2 category and KEGG pathways
m_t2g <- msigdbr(species = "Homo sapiens", category = "C2", subcategory = "KEGG") %>% 
  dplyr::select(gs_name, entrez_gene, gene_symbol)
head(m_t2g)

# This function will perform the GSEA
set.seed(2635847) # You need to set seed to get the same results. The method's test is based on permutation
em2 <- GSEA(geneList3, TERM2GENE = m_t2g, pvalueCutoff = 0.05) 
head(em2)

# Plot the top enriched pathways, annotating the identified proteins' location in the ranking 
gsea_p = gseaplot2(em2, geneSetID = 1:5, subplots = 1:2,
                   color=c("#0072B2", "#D95F02", "#E69F00", "#CC79A7", "#009E73"))
print(gsea_p) + 
  geom_vline(xintercept=(length(geneList3)/2), size=0.75, linetype=1)
ggsave(paste("gsea_kegg_signatures.pdf", sep=""), dpi=300, width=8, height=6)

# Select pathways that you want to plot with a ridgeplot
# If you want to select based on NES quantiles (e.g. for too many hits in REACTOME analyses)
# a = quantile(em2@result$NES, 0.9)
# b = quantile(em2@result$NES, 0.1)
# sel = c(which(em2@result$NES>a), which(em2@result$NES<b))

pem = em2
chos = em2@result$ID#[sel]
ch = list()

for (i in chos) {
  ch[i] = list(c(grep(i, pem@result$ID)))
}

pm = pem
pm@result = pm@result[unlist(ch),]

r = ridgeplot(pm, showCategory = 40) + theme_classic(base_size=9)
r

ggsave("ridgeplot_kegg.pdf", r, width = 8, height = 4, units = 'in', dpi=600)

# Convert back ENREZ IDs to gene names and save the results
em2 <- setReadable(em2, org.Hs.eg.db, keyType = "ENTREZID")
# View(em2@result)
write.csv(em2@result, "msigdb_gsea_differentially_expressed_in_covid_KEGG_sign_5%_fdr.csv")

```

Code for the protein-protein correlation network
- The same code for the protein-protein correlation matrix has been used for correlating protein levels to variables from the clinical dataset. In that analysis we excluded the healthy controls. 
```{r protein_correlation_network, include=FALSE, echo=FALSE}

dfc = t(dfaf)
cor_proteins = cor(dfc, dfc, method="spearman", use="pairwise.complete.obs")

x = colnames(dfc)
filename = "COVID_protein_protein"
  
protein_corr = list()
corr_m = list()
p_val = list()
for (i in x) {
    protein_corr[[i]] = corAndPvalue(dfc[,i], dfc,
                                     method="spearman", "pairwise.complete.obs")
  corr_m[[i]] = unlist(protein_corr[[i]]$cor)
  p_val[[i]] = unlist(protein_corr[[i]]$p)
  
}
  
matrix_protein_corr = as.matrix(do.call(rbind, corr_m))
rownames(matrix_protein_corr) = colnames(matrix_protein_corr)
matrix_protein_p_val = as.matrix(do.call(rbind, p_val))  
  
matrix_protein_corr_cat <- ifelse(matrix_protein_corr>0, "Positive", "Negative")

# protch = "VIL1" # Choose a protein of interest
# View(matrix_protein_corr[,protch])
  
m = matrix_protein_corr
m[upper.tri(m, diag=T)] = NA
  
c = matrix_protein_corr_cat
c[upper.tri(c, diag=T)] = NA
  
p = matrix_protein_p_val
p[upper.tri(c, diag=T)] = NA
  
med <- round(median(m, na.rm = TRUE), 2)
  
adjPvalue <- matrix(p.adjust(p, 
    method = "fdr"), nrow=nrow(matrix_protein_p_val), ncol=ncol(matrix_protein_p_val))
colnames(adjPvalue) = colnames(matrix_protein_p_val)
rownames(adjPvalue) = colnames(matrix_protein_p_val)

write.csv(matrix_protein_corr, paste("protein_protein_correlations_", 
    filename, ".csv", sep = ""), quote = FALSE)
write.csv(matrix_protein_corr_cat, paste("protein_protein_correlations_categorical_", 
    filename, ".csv", sep = ""), quote = FALSE)
write.csv(adjPvalue, paste("protein_protein_correlations_adj_p_", 
    filename, ".csv", sep = ""), quote = FALSE)

```